DOpElib: Differential Equations and Optimization Environment; A Goal Oriented Software Library for Solving PDEs and Optimization Problems with PDEs

### 2012-12-05 Paper_May_2012/main.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
In this article, we describe the  
\textit{Differential Equations and Optimization Environment} (\dope{}).
\dope{} is a software library that provides a unified interface 
to high level algorithms such as time-stepping methods, nonlinear solvers and 
optimization routines. This structure ensures 
that, first of all, the user is only required to write those sections of code 
that are specific to 
the considered problem. Second, the exchange of parts of the used routines is possible 
with only a few lines of code to change instead of large reimplementations.
The article illustrates the design principles and various features
of \dope{} and provides some 
numerical results as demonstration for the versatility of the software.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{introduction}
Today, there exists a broad variety of software libraries
that provide numerical tools for solving partial differential
equations (PDEs). Prominent examples are 
\deal{} \citep{dealnew}, dune \citep{dune}, 
FEniCS \citep{fenics}
or commercial solvers as for instance ANSYS (with its Fluent package),
COMSOL, or simufact forming.
Finally, a lot of work is (still) done in Matlab.
In fact, these libraries provide very useful tools 
for a huge variety of applications.

@@

\newpage
\subsection{Compliance Minimization}
In this example, we consider the compliance minimization of a standard MBB-Beam (Messerschmidt-B\"olkow-Blohm-Beam), see, e.g.,~\cite{BendsoeSigmund:2003}.
The thickness variable is allowed to take intermediate value apart from $0$ and $1$,
i.e., we consider a variable thickness sheet. The discretization is done 
using $Q_2$ elements for the displacement and discontinuous $P_0$ elements for the 
thickness.
\begin{figure}
\centering
\includegraphics[width=0.75\textwidth, viewport=205 88 975 500, clip]{Pictures/MBB.png}
\caption{Thickness distribution for the MBB-beam minimum compliance problem.} 
\label{res:mbb}
\end{figure}
A typical thickness distribution is shown in Fig.~\ref{res:mbb}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
In this article, we describe the  
\textit{Differential Equations and Optimization Environment} (\dope{}).
\dope{} is a software library that provides a unified interface 
to high level algorithms such as time-stepping methods, nonlinear solvers and 
optimization routines. This structure ensures 
that, first of all, the user is only required to write those sections of code 
that are specific to 
the considered problem. Second, the exchange of parts of the used routines is possible 
with only a few lines of code to change instead of large reimplementations.
The article illustrates the design principles and various features
of \dope{} and provides some 
numerical results as demonstration for the versatility of the software.

\keywords{software environment \and goal oriented interface design \and numerical solution of optimization problems and partial differential equations}
\subclass{49-04 \and  35-04}
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{introduction}
Today, there exists a broad variety of software libraries
that provide numerical tools for solving partial differential
equations (PDEs). Prominent examples are 
\deal{} \citep{dealnew}, dune \citep{dune}, 
FEniCS \citep{fenics}
or commercial solvers as for instance ANSYS (with its Fluent package),
COMSOL, or simufact forming.
Finally, a lot of work is (still) done in Matlab.
In fact, these libraries provide very useful tools 
for a huge variety of applications.

@@

\newpage
\subsection{Compliance Minimization}
In this example, we consider the compliance minimization of a standard MBB-Beam 
(Messerschmidt-B\"olkow-Blohm-Beam), see, e.g.,~\cite{BendsoeSigmund:2003}. 
In order to enforce thickness values to be either $1$ or $0$ we employ 
the, so called, SIMP-model (Solid Isotropic Material with Penalization), 
see as well,~\cite{BendsoeSigmund:2003}.

If we pick the SIMP paramter $p=1$ 
the thickness variable is allowed to take intermediate value apart from $0$ and $1$,
i.e., we consider a variable thickness sheet. The discretization is done 
using $Q_2$ elements for the displacement and discontinuous $P_0$ elements for the 
thickness.
\begin{figure}
\centering
\hspace*{\fill}
\includegraphics[width=0.44\textwidth, viewport=205 88 975 500, clip]{Pictures/MBB.png}
\hspace{\fill}
\includegraphics[width=0.44\textwidth, viewport=205 88 975 500, clip]{Pictures/MBB-top.png}
\hspace{\fill}
\caption{Thickness distribution for the MBB-beam minimum compliance problem using SIMP parameter $p=1$ (left) and $p=4$ right.} 
\label{res:mbb}
\end{figure}
A typical thickness distribution is shown in the left image in Fig.~\ref{res:mbb}.
To obtain thickness values that are, mostly, either $1$ or $0$ we need 
to pick a larger value for $p$. The distribution given in the right image in Fig.~\ref{res:mbb}
has been obtained using $p=4$.
